A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-20 paper.tex
Stab at getting eARG section finished

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the \emph{outcome} of such a random process or an inference
of the genetic genealogy of a sample of genome sequences.
However, the distinction between process and structure is not clear cut
and subfields use the term
differently (see Appendix XXX).
%% Sentence deleted: point made already
% ARG inference was for many years synonymous [FIXME: this isn't really true,
% what about parsimony ARGs?]
% with estimating realisations
% of the coalescent with recombination (with, it must be said, limited success),
% and the distinction between process and realisation was not especially important.

@@

\section*{Event ARGs}\label{eARG}
Genome ARGs are concerned with
describing the details of genetic inheritance
between specific pairs of genomes.
In contrast, ARGs are classically seem as being defined
by the evolutionary \emph{events} in the history of a sample.
The difference between gARGs and this prevalent view of
what defines an ARGs is subtle and the distinctions sometimes porous.
It is therefore useful to first survey some representative
definitions of
an ARG from the literature, to allow us to capture the common features
of these definitions
(see Appendix XXX for a full review).

@@

% Papers that do not discuss how positions work:
% - Mathieson and Scally: no discussion of how it works concretely (but quite
%   close to gARG, in some ways)
% TODO list out some of these as examples of where people don't bother
% getting into this important detail?
The common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``Event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occured,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use a common convention here
that the order of parents
% TODO list papers where people use ordering? ARGweaver I think?
is significant,
but additional information is sometimes incorporated to
distinguish left and right parents
\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

Beyone the interpretation of nodes as representing events or
genomes (which may reasonably be seen as somewhat philosophical),
the essential difference between a gARG and a classical eARG
[JK WIP from here:]
is that eARGs define the

carry positional information on their nodes, rather than
storing genome positions as edge annotations.

the regions of the genome which are transmitted are usually
specified by associating recombination nodes with a breakpoint.
In other words, eARGs
carry positional information on their nodes, rather than
storing genome positions as edge annotations.

To recover the local tree at position $x$,
requires traversing the graph rootwards from the leaves.
When we encounter a recombination node during traversal
if $x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent. This means the order in which parents are
listed is important.

% Jere & Yan: we thought about going into how simulation is difficult
% with just one recombination node due to needing an order before
% one is available during runtime. But since that can be solved
% fairly trivially with two recombination nodes, we decided it was
% more fiddly than it's worth.
% However, see https://github.com/tskit-dev/what-is-an-arg-paper/discussions/41#discussioncomment-2035609 for additional complexities

@@

\node [anchor=north west] at (8,6) {B};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{c|c|l}
% \multicolumn{2}{c}{Breakpoints}\\
Node & Breakpoint & Parents\\
\hline
$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{g}$ & $\varnothing$ & []\\
\end{tabular}};

% \node [anchor=north west] at (20,6) {C};
% \node [nodelabel,anchor=north west] at ($(10,5)$) {
% \begin{tabular}{ll}
% % Edges & Intervals\\
% \hline
% $(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
% $(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
% $(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
% $(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
% $(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
% $(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
% $(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
% \end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A classical event ARG (left) and a corresponding
encoding of the information (right).
Recombination nodes (e.g.\ \noderef{d}) have a non-null breakpoint
and two parents.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
}
\end{figure}

% FIXME this isn't worth getting into
Storing only breakpoint positions, rather than transmitted regions,
makes it hard to perform genetic analyses (such as extracting local trees)
without traversing the entire graph.
% "For this reason"?
%% Is the above true?? It the reason for this to allow efficient tree extraction? Are there other examples where encodings have been
%% built on top of an eARG-like node annotation scheme (ARGON?) and made efficient somehow?

[What is the ARG definition provided by the parsimony ARG literature?]

%[What is Gusfields ARG definition? Presumably it's still an eARG, but with a
%quirky dependence on SNP data?] DONE

Encoding genome positions on edges, rather than breakpoints on the nodes of an eARG, has important implications for efficiency, especially in forward-time simulations (see XXX)

% \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
\section*{Prospective vs retrospective ARGs}
ARGs are usually thought of as a retrospective structure, in which we
reason about the genomes that are genetic ancestors of some set of
sample genomes. Figure~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
Genetic transmission must have occurred between (e.g.)
$I_8$ and her daughter $I_6$, but this is omitted because
% this is probably not true, given we know the complete pedigree?
it is not relevant to the genetic ancestry of the samples

The gARG representation, on the other hand,
can be either retrospective or \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
% Wilder: finds the cell-to-cell confusing. Maybe drop?
number of generations (ultimately from cell to cell),
and are annotated
with the genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these  inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
direct genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{minichiello2006mapping} [more]
which we avoid due the potential confusion with statistical
marginal distrubtions [FIXME Gregor, what's the right wording here?]

>>>>

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the \emph{outcome} of such a random process or an inference
of the genetic genealogy of a sample of genome sequences.
However, the distinction between process and structure is not clear cut
and subfields use the term
differently (see Appendix XXX).
%% Sentence deleted: point made already
% ARG inference was for many years synonymous [FIXME: this isn't really true,
% what about parsimony ARGs?]
% with estimating realisations
% of the coalescent with recombination (with, it must be said, limited success),
% and the distinction between process and realisation was not especially important.

@@

\section*{Event ARGs}\label{eARG}
Genome ARGs are concerned with
describing the details of genetic inheritance
between specific pairs of genomes.
In contrast, ARGs are classically seem defined
by the evolutionary \emph{events} in the history of a sample.
The difference between gARGs and this prevalent view of
what defines an ARG is subtle and the distinctions sometimes porous.
It is therefore useful to first survey some representative
definitions of
an ARG from the literature, to allow us to capture the common features
of these definitions
(see Appendix XXX for a full review).

@@

% Papers that do not discuss how positions work:
% - Mathieson and Scally: no discussion of how it works concretely (but quite
%   close to gARG, in some ways)
% TODO list out some of these as examples of where people don't bother
% getting into this important detail?
The common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``Event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occured,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Figure~\ref{fig-arg-data-structure}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

@@

\node [anchor=north west] at (8,6) {C};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{c|c|l}
% \multicolumn{2}{c}{Breakpoints}\\
Node & Breakpoint & Parents\\
\hline
$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{g}$ & $\varnothing$ & []\\
\end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A classical event ARG. (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$.
(C) A concrete encoding of the information (right).
Recombination nodes have a non-null breakpoint
and two parents.
}
\end{figure}

Whether we regard nodes as events or genomes is something of a
philosophical question, and in most senses the two are equivalent
(but see the discussion on ``unary nodes'' in section XXX).
From a concrete data-structure encoding perspective, the only real
difference between an eARG and a gARG is that we store recombination
\emph{breakpoints} at nodes of out-degree two (recombination nodes)
in an eARG and we store \emph{inheritance intervals} on the edges
of a gARG. A key advantage of associating inheritance intervals with
edges in a gARG is that it allows us to describe evolutionary events other
than simple crossovers, such as multiple recombinations and
gene-conversions. By explicitly stating the exact regions of genome
that are transmitted along an edge, we generalise the representation
to encompass any process that leads to the differential inheritance of
genetic material along the genome.

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks. The vitally important and fundamental distinction between
a gARG and the classical eARG perspective is that gARGs do not
\emph{require} us to posit explicit recombination events in order
to describe a dataset, but
instead provide a means of encoding the effects of
multiple events, and to systematically describe the uncertainty around them
(see Section XXX).
As we see in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has many advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.

\section*{Prospective vs retrospective ARGs}
Event ARGs, as defined in the previous section as a way of encompassing
the range of definitions provided in the literature, are
inherently \emph{retrospective}.
Event ARGs are defined in terms
of the events that occur in the history of some set of sampled
sequences.
For example, Figure~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
Genetic transmission must have occurred between (e.g.)
$I_8$ and her daughter $I_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
it is not relevant to the genetic ancestry of the samples.

Genome ARGs, on the hand,
can be either retrospective or \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
% Wilder: finds the cell-to-cell confusing. Maybe drop?
number of generations (ultimately from cell to cell),
and are annotated
with the genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these  inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
direct genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

@@

\section*{Equivalance of ARGs and the set of local trees}
An ARG is sometimes referred to as being equivalent to a set of
local trees along the genome [TODO citation], but the connection
between the two is subtle.
In order for the mapping to be one-to-one, so that the original
gARG can be exactly recovered from a set of local trees and
their genome coordinates,
some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{minichiello2006mapping,griffiths1996ancestral} [more]
which we avoid due the potential confusion with statistical
marginal distrubtions [FIXME Gregor, what's the right wording here?]
